Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations

### 2017-02-21 ParisEM.tex
Final revision #4

\phi(x) =\|\alpha(x)\|^2  + \triangle A(x)\eqsp,

@@

In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50). The particle approximation of $\mathcal{Q}(\theta,\theta)$ is computed using each algorithm. An example of 200 estimates for a sampled set of observations is displayed in Figure~\ref{fig:obs:SINE}.

This estimation is performed 300 times with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.  Moreover, the E step boils down to the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.  $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}. This experiment was reproduced on a hundred different simulated data, in order study the distribution of the empirical absolute relative bias and coefficient of variations for all methods. The results are shown on Figure \ref{fig:mult:SINE}.

The GRand PaRIS algorithm outperforms the fixed lag methods for any value of the lag as the bias is the lowest (it is already negligible for $N=400$,) and with a lower variance than fixed lag estimates with negligible bias (i.e., in this case, lags larger than 10). Small lags lead to strongly biased estimates for the fixed lag method, and unbiased estimates are at the cost of a large variance. It is worth noting here that the lag for which the bias is small is model dependent. 
%To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated. Instead of the Ozaki discretization, the following bridges process are considered:
%\begin{enumerate}[i)]
%\item a bridge with constant drift as proposed in \cite{giesecke:schwenkler:2016}.
%\item a guided bridge process, see \cite{}.
%\end{enumerate}
%\paragraph{Choice of $\mu$}
%A simple choice is to set $\mu$ as a Poisson distribution with mean $\Lambda_{\theta}t$ leading to the first estimator GPE-1:
%\begin{multline*}
%q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) =  \widetilde{q}^{t}_{\theta}(x,y)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\Lambda_{\theta}^{-\kappa}\\
%\times \prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
%\end{multline*}

@@

The process solution to \eqref{eq:Lamp:LG} is observed regularly at times $t_0=0,\dots,t_{50}=100$ through the observation process $(Y_k)_{0\le k\le 50}$ defined as:
\begin{equation}
Y_k = X_k + \varepsilon_k\label{eq:obs:LG}\eqsp,
\end{equation}
where the $(\varepsilon_k)_{0\le k \le 50}$ are i.i.d. $ \mathcal{N}(0,\sigma^2_{obs})$.
The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the resulting GRand PaRIS algorithm has therefore a linear complexity. The intermediate quantity of the EM algorithm is evaluated  as for the SINE model.

@@

\caption{\textit{Left: }Observations of the process $X$ solution to the SDE \eqref{eq:Lamp:SINE} (black) and observations $Y$ satisfying equation \eqref{eq:obs:SINE} (white) at times $t_0=0,\dots,t_{100}=50$. \textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 300 replicates. The whiskers represent the extent of the 95\% central values. The dot represents the empirical mean over the 300 replicates. The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=10000$ particles.}

@@

\caption{Evolution of the empirical relative bias for each method in the case of the log-growth model. 100 replcations of the experience showed on Figure \ref{fig:res:SINE} are made. The empirical abolute relative bias is computed for each one. The whiskers show the 95\% central values and the dot shows the empirical median. All estimations are performed with an equal computation time.}

@@

\caption{\textit{Left: }Observations of the process $X$ solution to the SDE \eqref{eq:Lamp:LG} (black) and observations $Y$ satisfying equation \eqref{eq:obs:LG} (white) at times $t_0=0,\dots,t_{100}=50$. \textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 300 replicates. The whiskers represent the extent of the 95\% central values. The dot represents the empirical mean over the 300 replicates. The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=10000$ particles.}

@@

\caption{Empirical distribution of the absolute relative bias and te coefficient of variation of each estimation method. 100 replications of the experience showed on figure \ref{fig:res:LG} are made. The whiskers show the 95\% central values and the dot shows the empirical median. All estimations are performed with an equal computation time.}

>>>>

\phi(x) =\left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2\eqsp,

@@

In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50). The particle approximation of $\mathcal{Q}(\theta,\theta)$ is computed using each algorithm, see Figure~\ref{fig:res:SINE}. This estimation is performed 200 times with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.  Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.  $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.

\textcolor{red}{Detailler plus clairement ce que dit la figure 2 ici (donner clairement les estimateurs voire m\^eme les formules pour le biais et la variance relatifs)}
This experiment was reproduced on a hundred different simulated data, in order study the distribution of the empirical absolute relative bias and coefficient of variations for all methods. The results are shown on Figure \ref{fig:mult:SINE}.

The GRand PaRIS algorithm outperforms the fixed lag methods for any value of the lag as the bias is the lowest (it is already negligible for $N=400$) and with a lower variance than fixed lag estimates with negligible bias (i.e., in this case, lags larger than 10). Small lags lead to strongly biased estimates for the fixed lag method, and unbiased estimates are at the cost of a large variance. It is worth noting here that the lag for which the bias is small is model dependent. 
%To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated. Instead of the Ozaki discretization, the following bridges process are considered:
%\begin{enumerate}[i)]
%\item a bridge with constant drift as proposed in \cite{giesecke:schwenkler:2016}.
%\item a guided bridge process, see \cite{}.
%\end{enumerate}
%\paragraph{Choice of $\mu$}
%A simple choice is to set $\mu$ as a Poisson distribution with mean $\Lambda_{\theta}t$ leading to the first estimator GPE-1:
%\begin{multline*}
%q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) =  \widetilde{q}^{t}_{\theta}(x,y)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\Lambda_{\theta}^{-\kappa}\\
%\times \prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
%\end{multline*}

@@

The process solution to \eqref{eq:Lamp:LG} is observed regularly at times $t_0=0,\dots,t_{50}=100$ through the observation process $(Y_k)_{0\le k\le 50}$ defined as:
\begin{equation}
Y_k = X_k + \varepsilon_k\label{eq:obs:LG}\eqsp,
\end{equation}
where the $(\varepsilon_k)_{0\le k \le 50}$ are i.i.d. $ \mathcal{N}(0,\sigma^2_{obs})$.
The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the resulting GRand PaRIS algorithm has therefore a linear complexity. The intermediate quantity of the EM algorithm is evaluated  as for the SINE model, see Figures~\ref{fig:res:LG} and~\ref{fig:mult:LG}.

@@

\caption{\textit{Left: }Process $X$ solution to the SDE \eqref{eq:Lamp:SINE} (black) and observations $Y$ satisfying equation \eqref{eq:obs:SINE} (white) at times $t_0=0,\dots,t_{100}=50$. \textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 200 replicates. The whiskers represent the extent of the 95\% central values. The dot represents the empirical mean over the 200 replicates. The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}

@@

\caption{Evolution of the empirical relative bias for each method in the case of the log-growth model. 100 replications of the experience showed on Figure \ref{fig:res:SINE} are made. The empirical abolute relative bias is computed for each one. The whiskers show the 95\% central values and the dot shows the empirical median. All estimations are performed with an equal computation time.}

@@

\caption{\textit{Left: }Process $X$ solution to the SDE \eqref{eq:Lamp:LG} (black) and observations $Y$ satisfying equation \eqref{eq:obs:LG} (white) at times $t_0=0,\dots,t_{100}=50$. \textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 200 replicates. The whiskers represent the extent of the 95\% central values. The dot represents the empirical mean over the 200 replicates. The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}

@@

\caption{Empirical distribution of the absolute relative bias and the coefficient of variation of each estimation method. 100 replications of the experience showed on figure \ref{fig:res:LG} are made. The whiskers show the 95\% central values and the dot shows the empirical median. All estimations are performed with an equal computation time.}
